Addition and removal energies of circular quantum dots

### 2017-06-13 Manuscript/paper.tex
small revisions

Even though the model system contains five parameters $(N, m, \omega, e, \epsilon)$, many of them are redundant.  With an appropriate choice of units, the number of parameters in the system can be reduced to just two.  For rest of this paper, we shall use atomic units where $\hbar$, $m$, $e$ and $4 \pi \epsilon$ are all set to one.  The unit Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2 &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}.
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

@@

\begin{figure}
  \includegraphics{figures/fig-shell-structure-v2.pdf}
  \caption{The 42 lowest single-particle states (the first 5 shells) in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  Within each column, the principal quantum number $n$ increases as one traverses upward.}
  \label{fig:shell-structure}
\end{figure}

@@

Since electrons are spin-$\frac{1}{2}$ fermions, they can occupy either of the two possible spin states $\chi_{-\frac{1}{2}}$ or $\chi_{+\frac{1}{2}}$.  Thus, every single-particle basis state $|n m_\ell m_{\mathrm{s}}\rangle$ contains both a spatial component \eqref{eq:fockdarwin} and a spin component,
\begin{align} \label{eq:singleparticlestate}
  \langle \rho \varphi m_{\mathrm{s}}' |n m_\ell m_{\mathrm{s}}\rangle \equiv F_{n m_\ell}(\rho, \varphi) \delta_{m_{\mathrm{s}}^{} m_{\mathrm{s}}'}.
\end{align}
Here we have introduced spin projection $m_{\mathrm{s}} \in \bigl\{-\frac{1}{2}, +\frac{1}{2}\bigr\}$ as the third quantum number and $\delta_{p q}$ to denote the Kronecker delta..

@@

The energy of a Slater determinant is given by a sum over the occupied single-particle energies,
\begin{align*}
  \hat{H}^{[1]} | p_1 \ldots p_N \rangle = (\varepsilon_{p_1} + \cdots + \varepsilon_{p_N}) | p_1 \ldots p_N \rangle
\end{align*}
It is possible for Slater determinants occupying different sets of single-particle states to be degenerate in energy.  This is possible even for ground states of $\hat{H}^{[1]}$.  However, if the number of particles $N$ satisfies $N = K_{\mathrm{F}} (K_{\mathrm{F}} + 1)$ for some nonnegative integer $K_{\mathrm{F}}$, there would be just enough particles to form a closed-shell Slater determinant, leading to a unique, well-isolated ground state.  The values of $N$ at which this occurs are said to be \textit{magic numbers}.  We call $K_{\mathrm{F}}$ the \textit{number of filled shells} (or ``Fermi level'').  In particular, a single-particle state is occupied in the ground state Slater determinant if and only $k < K_{\mathrm{F}}$, where $k$ is the shell index of the single-particle state as defined in \eqref{eq:shell_index}.

@@

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.\ \eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{}.
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.

The complete Hamiltonian is given by the sum
\begin{align}
  \hat{H} = \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r. \label{eq:vacuumhamiltonian}
\end{align}
Due to the antisymmetry of Slater determinants, we have the property that exchanging either $p$ with $q$ or $r$ with $s$ reverses the sign of $\langle p q | \hat{H}^{[2]} | r s \rangle$, leading to a 4-fold symmetry that must be canceled out by the $1/4$ factor in front of the summation in \eqref{eq:vacuumhamiltonian}.  For this reason, the two-body matrix element $\langle p q | \hat{H}^{[2]} | r s \rangle$ is commonly referred to as the \emph{antisymmetrized} matrix element.  This is not to be confused with the (non-antisymmetric) interaction integral,
\begin{align} \label{eq:interactionintegral}
  v_{p_1 p_2 p_3 p_4} \equiv \int \langle p_1 | \bm{r} \rangle \langle p_2 | \bm{r}' \rangle v(\bm r, \bm r') \langle \bm{r} | p_3 \rangle \langle \bm{r}' | p_4 \rangle \D \bm{r} \D \bm{r}'.
\end{align}
This object only has a two-fold symmetry: $v_{p_1 p_2 p_3 p_4} = v_{p_2 p_1 p_4 p_3}$.  The antisymmetrized matrix elements are related to the interaction integral via
\begin{align} \label{eq:antisymmetricmatrixelement}
  \langle p q | \hat{H}^{[2]} | r s \rangle = v_{p q r s} - v_{p q s r}
\end{align}

@@

One of the simplest numerical corrections that could be applied to the initial guess is that of the Hartree--Fock (HF) method.  Using the variational principle, one can compute an approximate ground state $\Phi$ by minimizing the energy expectation value
\begin{align} \label{eq:hfexpectation}
  E_{\Phi} \equiv \langle \Phi | \hat H | \Phi \rangle
\end{align}
with respect to $|\Phi\rangle$, subject to the restriction that $|\Phi\rangle$ remains a single Slater determinant constructed from an unknown single-particle basis.  The restriction is what enables the simplicity and efficiency of this method.  We shall denote each unknown single-particle state $|q'\rangle$ by a primed label $q'$.

@@

The goal then is to find the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, now given by
\begin{align}
  E_{\Phi} &= \sum_i H^{[1] \prime}_{i i} + \sum_{i j} H^{[2] \prime}_{i j i j} \label{eq:hfenergy},
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{[1] \prime}_{p q} &\equiv \sum_{r s} C_{r p}^* \langle r | \hat{H}^{[1]} | s \rangle C_{s q}^{} \label{eq:hftransform1} \\
  H^{[2] \prime}_{p q r s} &\equiv \sum_{t u v w} C_{t p}^* C_{u q}^* \langle t u | \hat{H}^{[2]} | v w \rangle C_{v r}^{} C_{w s}^{}, \label{eq:hftransform2}
\end{align}

@@

The second-quantized operators in Eq.\ \eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1 i_2 \ldots i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label the occupied states.  One can define a set of \textit{quasiparticle field operators} $\hat b$ and $\hat b^\dagger$ that obey the same anticommutation relations as the $\hat a$ operators.  The normal ordering of the quasiparticle field operators is straightforward to compute.  What is more interesting is the normal ordering of the physical field operators $\hat a$ and $\hat a^\dagger$ with respect to $|\Phi\rangle$:
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &= \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p q}^{}.
\end{align*}
and $n_p$ is the \textit{occupancy} of the single-particle state $p$, defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state of $\Phi$}, \\
    0 & \text{if $p$ is an unoccupied state of $\Phi$}.
  \end{cases}
\end{align}

The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix $F_{p q}$.  The operator $\hat H$ in Eq.\ \eqref{eq:normordhamiltonian} is completely equivalent to $\hat H$ in Eq.\ \eqref{eq:vacuumhamiltonian}.  The only difference is that the meaning of a ``$k$-body operator'' has been redefined with respect to the reference state $\Phi$ rather than vacuum state $\ket{}$, causing matrix elements to be reshuffled among the $k$-body operators.  Where this makes a critical difference is when the operator expressions are \emph{truncated}, i.e.\ some $k$-body operators may be discarded from the computation for efficiency reasons.  Normal-ordering can drastically alter the importance of the $k$-body operators with respect to each other.

Higher-body operators arise from integrating the flow equations (\ref{eq:imsrgode}), which is one of the major challenges of the SRG method.  With each evaluation of the commutator, the Hamiltonian gains terms of higher order, and these induced contributions will in subsequent integration steps feed back into terms of lower order.  Thus, the higher-body contributions are not irrelevant to the final solution even if only the ground state energy (zero-body component) is of interest.

@@

By expanding the commutator in Eq.\ \eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H) \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \antisymm_{A B} D^{[k]}(A, B), \label{eq:commutmxe} \\
  \antisymm_{x y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr),
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p q}(A, B)
  &\equiv
    \sum_r A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right),
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    4 \antisymm_{p q} \antisymm_{r s} \sum_{i a} A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    + \frac{1}{2} \sum_{i j} A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + 2 \sum_t \left(
    \antisymm_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \antisymm_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}
    \right).
    \label{eq:flow2}
\end{align}

\begin{figure}
\includegraphics{fig-diagrams-imsrg.pdf}
\caption{Diagrammatic representation of $D^{[k]}(\circ, \bullet)$ of Eq.\ \eqref{eq:flow0}, Eq.\ \eqref{eq:flow1}, and Eq.\ \eqref{eq:flow2}, where open circles represent $A$ and filled circles represent $B$.  The two-body vertices are implicitly antisymmetric (i.e.\ \textit{Hugenholtz diagrams} \cite{HUGENHOLTZ1957481}, as described in \cite{shavitt2009many} \S 4.4.3).  Internal directed lines (bound/dummy indices) that point upward denote summations over unoccupied states, whereas lines that point downward denote summations over occupied states.  External undirected lines denote unconstrained free indices.  In IM-SRG(2), the three-body diagrams are not included.}
\label{fig:diagrams-imsrg}
\end{figure}

@@

The White generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta^{\prime [1]}_{a i} &\equiv \frac{H^{[1]}_{a i}}{\tilde{\Delta}^{[1]}_{a i}} &
\eta^{\prime [2]}_{a b i j} &\equiv \frac{H^{[2]}_{a b i j}}{\tilde{\Delta}^{[2]}_{a b i j}}
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}^{[1]}_{a i} &\equiv \Delta_{a i} - H^{[2]}_{a i a i}, \notag \\
\tilde{\Delta}^{[2]}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H^{[1]}_{p_i p_i} -  H^{[1]}_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H^{[2]}_{a b a b} - H^{[2]}_{a i a i} - H^{[2]}_{b i b i} \notag \\
  &\qquad + H^{[2]}_{i j i j} - H^{[2]}_{a j a j} - H^{[2]}_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

\begin{figure}
  \includegraphics{fig-diagrams-ccsd.pdf}
  \caption{Diagrammatic representation of $\bar{H}$ of Eq.\ \eqref{eq:cc_heff1}, where open circles represent the excitation cluster operators $\hat{T}_1$ and $\hat{T}_2$, and filled circles represent the two-body interaction $\hat{H}^{[2]}$.  As before, the diagrams are implicitly antisymmetrized (Hugenholtz diagrams).  Lines connected to $\hat{T}$ are always directed upward because they represent an excitation operator while the directions of external lines connected to $\hat{H}^{[2]}$ are unconstrained.}
  \label{fig:diagrams-ccsd}
\end{figure}

@@

\begin{figure}
\includegraphics{fig-diagrams-sfe.pdf}
\caption{Diagrammatic form of the second- and third-order QDPT corrections.  The diagrams are implicitly antisymmetrized (Hugenholtz diagrams), but also have implicit denominators as with many-body perturbation theory.  When QDPT is performed on IM-SRG-evolved Hamiltonians, many of the diagrams vanish.  The remaining nonvanishing diagrams for addition energy are highlighted in blue and for removal energy are highlighted in red.}
\label{fig:diagrams-sfe}
\end{figure}

>>>>

Even though the model system contains five parameters $(N, m, \omega,
e, \epsilon)$, many of them are redundant.  With an appropriate choice
of units, the number of parameters in the system can be reduced to
just two.  For rest of this paper, we shall use atomic units where
$\hbar$, $m$, $e$ and $4 \pi \epsilon$ are all set to one.  The unit
Hartree $E_{\mathrm{h}}$ becomes the unit of energy and the Bohr
radius $a$ becomes the unit of length:
\begin{align*}
  E_{\mathrm{h}} &\equiv m \left(\frac{e^2}{4 \pi \epsilon \hbar}\right)^2, &
  a &\equiv \frac{4 \pi \epsilon \hbar^2}{m e^2}.
\end{align*}
This leaves us with $(N, \omega)$ as the only parameters necessary to specify the quantum dot system, with $\omega$ in units of $E_{\mathrm{h}} / \hbar$.

@@

\begin{figure}
  \includegraphics{figures/fig-shell-structure-v2.pdf}
  \caption{(Color online) The 42 lowest single-particle states (the first 5 shells) in the 2D harmonic oscillator basis.  Each box represents a single-particle state arranged by $m_\ell$, $m_{\mathrm{s}}$, and energy, and the up/down arrows indicate the spin of the states.  Within each column, the principal quantum number $n$ increases as one traverses upward.}
  \label{fig:shell-structure}
\end{figure}

@@

Since electrons are spin-$\frac{1}{2}$ fermions, they can occupy
either of the two possible spin states $\chi_{-\frac{1}{2}}$ or
$\chi_{+\frac{1}{2}}$.  Thus, every single-particle basis state $|n
m_\ell m_{\mathrm{s}}\rangle$ contains both a spatial
component \eqref{eq:fockdarwin} and a spin component,
\begin{align} \label{eq:singleparticlestate}
  \langle \rho \varphi m_{\mathrm{s}}' |n m_\ell m_{\mathrm{s}}\rangle \equiv F_{n m_\ell}(\rho, \varphi) \delta_{m_{\mathrm{s}}^{} m_{\mathrm{s}}'}.
\end{align}
Here we have introduced spin projection
$m_{\mathrm{s}} \in \bigl\{-\frac{1}{2}, +\frac{1}{2}\bigr\}$ as the
third quantum number and $\delta_{p q}$ to denote the Kronecker
delta.

@@

The energy of a Slater determinant is given by a sum over the occupied single-particle energies,
\begin{align*}
  \hat{H}^{[1]} | p_1 \ldots p_N \rangle = (\varepsilon_{p_1} + \cdots + \varepsilon_{p_N}) | p_1 \ldots p_N \rangle.
\end{align*}
It is possible for Slater determinants occupying different sets of
single-particle states to be degenerate in energy.  This is possible
even for ground states of $\hat{H}^{[1]}$.  However, if the number of
particles $N$ satisfies $N = K_{\mathrm{F}} (K_{\mathrm{F}} + 1)$ for
some nonnegative integer $K_{\mathrm{F}}$, there would be just enough
particles to form a closed-shell Slater determinant, leading to a
unique, well-isolated ground state.  The values of $N$ at which this
occurs are said to be \textit{magic numbers}.  We call
$K_{\mathrm{F}}$ the \textit{number of filled shells} (or ``Fermi
level'').  In particular, a single-particle state is occupied in the
ground state Slater determinant if and only $k < K_{\mathrm{F}}$,
where $k$ is the shell index of the single-particle state as defined
in \eqref{eq:shell_index}.

@@

In second quantization, the one- and two-body Hamiltonian operators $H^{[1]}$ and $\hat{H}^{[2]}$ from Eq.\ \eqref{eq:onetwobodyhamiltonian} can be rewritten as
\begin{align} \label{eq:second_quantized_hamiltonian}
  \hat H^{[1]} &= \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat a_p^\dagger \hat a_q^{}, &
  \hat{H}^{[2]} &= \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat a_p^\dagger \hat a_q^\dagger \hat a_s^{} \hat a_r^{},
\end{align}
where the quantities $\langle p | \hat{H}^{[1]} | q \rangle$ and $\langle p q | \hat{H}^{[2]} | r s \rangle$ are known as \textit{matrix elements} and fully characterize the two operators in the basis.

The complete Hamiltonian is given by the sum
\begin{align}
  \hat{H} = \sum_{p q} \langle p | \hat{H}^{[1]} | q \rangle \hat{a}_p^\dagger \hat{a}_q + \frac{1}{4} \sum_{p q r s} \langle p q | \hat{H}^{[2]} | r s \rangle \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r. \label{eq:vacuumhamiltonian}
\end{align}
Due to the antisymmetry of Slater determinants, we have the property
that exchanging either $p$ with $q$ or $r$ with $s$ reverses the sign
of $\langle p q | \hat{H}^{[2]} | r s \rangle$, leading to a four-fold
symmetry that must be canceled out by the $1/4$ factor in front of the
summation in \eqref{eq:vacuumhamiltonian}.  For this reason, the
two-body matrix element $\langle p q | \hat{H}^{[2]} | r s \rangle$ is
commonly referred to as the \emph{antisymmetrized} matrix element.
This is not to be confused with the (non-antisymmetric) interaction
integral,
\begin{align} \label{eq:interactionintegral}
  v_{p_1 p_2 p_3 p_4} \equiv \int \langle p_1 | \bm{r} \rangle \langle p_2 | \bm{r}' \rangle v(\bm r, \bm r') \langle \bm{r} | p_3 \rangle \langle \bm{r}' | p_4 \rangle \D \bm{r} \D \bm{r}'.
\end{align}
This object only has a two-fold symmetry: $v_{p_1 p_2 p_3 p_4} = v_{p_2 p_1 p_4 p_3}$.  The antisymmetrized matrix elements are related to the interaction integral via
\begin{align} \label{eq:antisymmetricmatrixelement}
  \langle p q | \hat{H}^{[2]} | r s \rangle = v_{p q r s} - v_{p q s r}.
\end{align}

@@

One of the simplest numerical corrections that could be applied to the
initial guess is that of the Hartree--Fock (HF) method.  Using the
variational principle, one can compute an approximate ground state
$\Phi$ by minimizing the energy expectation value
\begin{align} \label{eq:hfexpectation}
  E_{\Phi} \equiv \langle \Phi | \hat H | \Phi \rangle,
\end{align}
with respect to $|\Phi\rangle$, subject to the restriction that
$|\Phi\rangle$ remains a single Slater determinant constructed from an
unknown single-particle basis.  The restriction is what enables the
simplicity and efficiency of this method.  We shall denote each
unknown single-particle state $|q'\rangle$ by a primed label $q'$.

@@

The goal then is to find the coefficients $\bm C$ that minimize the energy $E_{\Phi}$, now given by
\begin{align}
  E_{\Phi} &= \sum_i H^{[1] \prime}_{i i} + \sum_{i j} H^{[2] \prime}_{i j i j} \label{eq:hfenergy},
\end{align}
where the transformed operators are defined as
\begin{align}
  H^{[1] \prime}_{p q} &\equiv \sum_{r s} C_{r p}^* \langle r | \hat{H}^{[1]} | s \rangle C_{s q}^{}, \label{eq:hftransform1} \\
  H^{[2] \prime}_{p q r s} &\equiv \sum_{t u v w} C_{t p}^* C_{u q}^* \langle t u | \hat{H}^{[2]} | v w \rangle C_{v r}^{} C_{w s}^{}, \label{eq:hftransform2}
\end{align}

@@

The second-quantized operators in Eq.\ \eqref{eq:second_quantized_hamiltonian} are constructed with respect to the vacuum state in the sense that their operator strings $\hat{a}_p^\dagger \hat{a}_q$ and $\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r$ both have expectation values of zero in vacuum:
\begin{align*}
&\bra{} \hat{a}_p^\dagger \hat{a}_q \ket{} = 0, &
&\bra{} \hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s \hat{a}_r \ket{} = 0.
\end{align*}
They are said to be \emph{normal-ordered} with respect to the vacuum state $\ket{}$.  It is possible to be normal-ordered with respect to a different state, however.  Suppose we have a Slater determinant $\Phi$ with $N$ occupied states:
\begin{align*}
  \ket{\Phi} \equiv \ket{i_1 i_2 \ldots i_N}
\end{align*}
where $\{i_1, i_2, \ldots, i_N\}$ is a set of $N$ indices that label
the occupied states.  We define then the normal ordering of the
physical field operators $\hat a$ and $\hat a^\dagger$ with respect to
$|\Phi\rangle$, that is
\begin{align*}
  \normord{\hat{a}_p^\dagger \hat{a}_q^{}} &= \hat{a}_p^\dagger \hat{a}_q^{} - n_p^{} \delta_{p q}^{}.
\end{align*}
where $n_p$ is the \textit{occupancy} of the single-particle state $p$, defined as
\begin{align}
  n_p \equiv \begin{cases}
    1 & \text{if $p$ is an occupied state of $\Phi$}, \\
    0 & \text{if $p$ is an unoccupied state of $\Phi$}.
  \end{cases}
\end{align}

The normal ordering notation thus provides a concise way of expressing the right-hand side.  In this new notation, we can rewrite the Hamiltonian $\hat H = \hat{H}^{[1]} + \hat{H}^{[2]}$ in its \emph{normal-ordered representation}:
\begin{align}
  \hat{H} = H^{[0]} + \sum_{p q} H^{[1]}_{p q} \normord{\hat{a}_p^\dagger \hat{a}_q^{}} + \frac{1}{4} \sum_{p q r s} H^{[2]}_{p q r s} \normord{\hat{a}_p^\dagger \hat{a}_q^\dagger \hat{a}_s^{} \hat{a}_r^{}},
  \label{eq:normordhamiltonian}
\end{align}
where
\begin{align*}
  H^{[0]} &\equiv \sum_i \langle i | \hat{H}^{[1]} | i \rangle + \frac{1}{2} \sum_{i j} \langle i j | \hat{H}^{[2]} | i j \rangle, \\
  H^{[1]}_{p q} &\equiv \langle p | \hat{H}^{[1]} | q \rangle + \sum_i \langle p i | \hat{H}^{[2]} | q i \rangle, \\
  H^{[2]}_{p q r s} &\equiv \langle p q | \hat{H}^{[2]} | r s \rangle. \\
\end{align*}
If $\Phi$ is a state optimized by the HF method, then $H^{[0]}$ is
simply the HF energy $E_\Phi$ and $H^{[1]}_{p q}$ is the Fock matrix
$F_{p q}$.  The operator $\hat H$ in
Eq.\ \eqref{eq:normordhamiltonian} is completely equivalent to $\hat
H$ in Eq.\ \eqref{eq:vacuumhamiltonian}.  The only difference is that
the meaning of a ``$k$-body operator'' has been redefined with respect
to the reference state $\Phi$ rather than the vacuum state $\ket{}$,
causing matrix elements to be reshuffled among the $k$-body operators.
Where this makes a critical difference is when the operator
expressions are \emph{truncated}, i.e.\ some $k$-body operators may be
discarded from the computation for efficiency reasons.
Normal-ordering can drastically alter the importance of the $k$-body
operators with respect to each other.

Higher-body operators arise from integrating the flow equations of Eq.~(\ref{eq:imsrgode}), 
which is one of the major challenges of the SRG
method.  With each evaluation of the commutator, the Hamiltonian gains
terms of higher order, and these induced contributions will in
subsequent integration steps feed back into terms of lower order.
Thus, the higher-body contributions are not irrelevant to the final
solution even if only the ground state energy (zero-body component) is
of interest.

@@

By expanding the commutator in Eq.\ \eqref{eq:imsrgode} and discarding the three-body term, we obtain the matrix-element form of the IM-SRG(2) flow equation:
\begin{align}
    \frac{\D H^{[k]}}{\D s} = C^{[k]}(\eta, H), \label{eq:flowmxe}
\end{align}
where
\begin{align}
  C^{[k]}(A, B) &\equiv 2 \antisymm_{A B} D^{[k]}(A, B), \label{eq:commutmxe} \\
  \antisymm_{x y} f(x, y) &\equiv \frac{1}{2} \bigl(f(x, y) - f(y, x)\bigr),
\end{align}
\begin{align}
  D^{[0]}(A, B)
  &\equiv
    \sum_{i a} A^{[1]}_{i a} B^{[1]}_{a i}
    + \frac{1}{4} \sum_{i j a b} A^{[2]}_{i j a b} B^{[2]}_{a b i j},
    \label{eq:flow0} \\
  D^{[1]}_{p q}(A, B)
  &\equiv
    \sum_r A^{[1]}_{p r} B^{[1]}_{r q}
    - \frac{1}{2} \sum_{i j a} A^{[2]}_{i j a q} B^{[2]}_{a p i j}
    + \frac{1}{2} \sum_{i a b} A^{[2]}_{i p a b} B^{[2]}_{a b i q}
    \notag \\
  &\hphantom{=}
    + \sum_{i a} \left(
    A^{[1]}_{i a} B^{[2]}_{a p i q}
    + A^{[2]}_{i p a q} B^{[1]}_{a i}
    \right),
    \label{eq:flow1} \\
  D^{[2]}_{p q r s}(A, B)
  &\equiv
    4 \antisymm_{p q} \antisymm_{r s} \sum_{i a} A^{[2]}_{i p a r} B^{[2]}_{a q i s}
    + \frac{1}{2} \sum_{i j} A^{[2]}_{i j r s} B^{[2]}_{p q i j}
    + \frac{1}{2} \sum_{a b} A^{[2]}_{p q a b} B^{[2]}_{a b r s}
    \notag \\
  &\hphantom{=}
    + 2 \sum_t \left(
    \antisymm_{p q} A^{[1]}_{q t} B^{[2]}_{p t r s}
    + \antisymm_{r s} A^{[2]}_{p q r t} B^{[1]}_{t s}
    \right).
    \label{eq:flow2}
\end{align}

\begin{figure}
\includegraphics{fig-diagrams-imsrg.pdf}
\caption{(Color online) Diagrammatic representation of $D^{[k]}(\circ, \bullet)$ of Eq.\ \eqref{eq:flow0}, Eq.\ \eqref{eq:flow1}, and Eq.\ \eqref{eq:flow2}, where open circles represent $A$ and filled circles represent $B$.  The two-body vertices are implicitly antisymmetric (i.e.\ \textit{Hugenholtz diagrams} \cite{HUGENHOLTZ1957481}, as described in \cite{shavitt2009many} \S 4.4.3).  Internal directed lines (bound/dummy indices) that point upward denote summations over unoccupied states, whereas lines that point downward denote summations over occupied states.  External undirected lines denote unconstrained free indices.  In IM-SRG(2), the three-body diagrams are not included.}
\label{fig:diagrams-imsrg}
\end{figure}

@@

The White generator is explicitly constructed the following way \cite{PhysRevLett.106.222502,White:cond-mat0201346}.  Let
\begin{align*}
\hat{\eta}^{\text{Wh}} &\equiv \hat{\eta}' - \hat{\eta}'{}^\dagger,
\end{align*}
where
\begin{align*}
\eta^{\prime [1]}_{a i} &\equiv \frac{H^{[1]}_{a i}}{\tilde{\Delta}^{[1]}_{a i}}, &
\eta^{\prime [2]}_{a b i j} &\equiv \frac{H^{[2]}_{a b i j}}{\tilde{\Delta}^{[2]}_{a b i j}},
\end{align*}
and the Epstein--Nesbet energy denominators $\tilde{\Delta}$ \cite{shavitt2009many} are defined as
\begin{align}
\tilde{\Delta}^{[1]}_{a i} &\equiv \Delta_{a i} - H^{[2]}_{a i a i}, \notag \\
\tilde{\Delta}^{[2]}_{a b i j} &\equiv \Delta_{a b i j} + w_{a b i j}, \notag \\
\Delta_{p_1 \ldots p_k q_1 \ldots q_k} &\equiv \sum_{i = 1}^k (H^{[1]}_{p_i p_i} -  H^{[1]}_{q_i q_i}), \label{eq:moellerplessetdenominator} \\
w_{a b i j}
  &\equiv H^{[2]}_{a b a b} - H^{[2]}_{a i a i} - H^{[2]}_{b i b i} \notag \\
  &\qquad + H^{[2]}_{i j i j} - H^{[2]}_{a j a j} - H^{[2]}_{b j b j}. \notag
\end{align}
It is also possible \cite{IMSRG} to use $\Delta$ (also defined above), the same energy denominators from M\o ller--Plesset perturbation theory, which leads to a slightly different variant of the White generator.  For our calculations, we use exclusively Epstein--Nesbet denominators.

@@

\begin{figure}
  \includegraphics{fig-diagrams-ccsd.pdf}
  \caption{(Color online) Diagrammatic representation of $\bar{H}$ of Eq.\ \eqref{eq:cc_heff1}, where open circles represent the excitation cluster operators $\hat{T}_1$ and $\hat{T}_2$, and filled circles represent the two-body interaction $\hat{H}^{[2]}$.  As before, the diagrams are implicitly antisymmetrized (Hugenholtz diagrams).  Lines connected to $\hat{T}$ are always directed upward because they represent an excitation operator while the directions of external lines connected to $\hat{H}^{[2]}$ are unconstrained.}
  \label{fig:diagrams-ccsd}
\end{figure}

@@

\begin{figure}
\includegraphics{fig-diagrams-sfe.pdf}
\caption{(Color online) Diagrammatic form of the second- and third-order QDPT corrections.  The diagrams are implicitly antisymmetrized (Hugenholtz diagrams), but also have implicit denominators as with many-body perturbation theory.  When QDPT is performed on IM-SRG-evolved Hamiltonians, many of the diagrams vanish.  The remaining nonvanishing diagrams for addition energy are highlighted in blue and for removal energy are highlighted in red.}
\label{fig:diagrams-sfe}
\end{figure}
